---
title: "1075 Publication Plots"
author: "Matthew Allen"
date: "2023-12-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Dependencies, Read Data, Create Objects
```{r, results='hide', message=FALSE, warning=FALSE}

    # Install and load required packages
    packages_to_install <- c("ggpubr", "gridExtra", "ggplot2", "knitr", "outliers", "dplyr", "readr")
    
    if (length(setdiff(packages_to_install, rownames(installed.packages()))) > 0) {
      install.packages(setdiff(packages_to_install, rownames(installed.packages())))
    }
    
    library(gridExtra)
    library(ggpubr)
    library(knitr)
    library(ggplot2)
    library(outliers)
    library(dplyr)

# Here is a custom delta notation object
  oxydelt <- expression("δ"^18 * "O‰ (VSMOW)")
  
# Read in raw data
    V1075raw_githubURL <- "https://github.com/mattgeo1990/1075_Vertebrate_d18Op/blob/main/Data/V1075_PhosphateData_8-18-23_copy.csv"
    V107_raw <- read.csv(V1075raw_githubURL)
  
# Read in cleaned data (include link to script that cleans the data?)
  # Call V1075_cl from GitHub 
  # cleaning script is here: https://github.com/mattgeo1990/1075_Vertebrate_d18Op/blob/main/Code/V1075_d18O.R
      V1075cl_githubURL <- "https://github.com/mattgeo1990/1075_Vertebrate_d18Op/blob/main/Data/V1075_cl.csv"
      V1075_cl <- read.csv(V1075cl_githubURL)

```

## Plot stripchart

```{r, results='hide', message=FALSE, warning=FALSE}
# d18O by Eco_type
  ggplot(V1075_cl, aes(x=reorder(Eco, d18O..VSMOW., na.rm = TRUE), y=d18O..VSMOW.))+
    theme(panel.background = element_rect(color = "black", size = 2, fill = NA)) +
    geom_jitter(position=position_jitter(0), shape = 1, size = 3, stroke = 1) +
    theme(panel.grid.major.x = element_line(color = "black", size = 0.1)) +
    theme(axis.text.x = element_text(size = 15, angle = -45, hjust = 0, face = "bold")) +
    theme(axis.text.y = element_text(size = 15, hjust = 0, color = "black")) +
    theme(axis.ticks.x = element_line())+
    theme(panel.grid.major.y = element_blank())+
    theme(panel.grid.minor.y = element_blank())+
    theme(plot.margin = margin(20, 40, 20, 20, "pt")) +
    theme(axis.title=element_text(size=20,face="bold")) +
    scale_y_continuous(breaks = seq(10, 25, by = 2)) +
    labs(x="", y = oxydelt)

# d18O by Taxon
  ggplot(V1075_cl, aes(x=reorder(Taxon, d18O..VSMOW., na.rm = TRUE), y=d18O..VSMOW.))+
      theme(panel.background = element_rect(color = "black", size = 2, fill = NA)) +
      geom_jitter(position=position_jitter(0), shape = 1, size = 3, stroke = 1) +
      theme(panel.grid.major.x = element_line(color = "black", size = 0.1)) +
      theme(axis.text.x = element_text(size = 15, angle = -45, hjust = 0, face = "bold")) +
      theme(axis.text.y = element_text(size = 15, hjust = 0, color = "black")) +
      theme(axis.ticks.x = element_line())+
      theme(panel.grid.major.y = element_blank())+
      theme(panel.grid.minor.y = element_blank())+
      theme(plot.margin = margin(20, 40, 20, 20, "pt")) +
      theme(axis.title=element_text(size=20,face="bold")) +
      scale_y_continuous(breaks = seq(10, 25, by = 2)) +
      labs(x="", y = oxydelt)
```

## Plot Latitudinal Gradients

#### Need to sort this out. Want to use base R or ggplot? Include error bars. What part of the Cretaceous do these models represent? What time slice should I include? Just Albian?

#### Plot 1: Temps vs latitude
  * Data to include:
    - 1075 Dual-taxon temps
    - 1075 EECM
    - Kate's Antler's phyllosilicate data
    - CMF temps
    
#### Plot 2: d18Omw vs latitude
 * Data to include:
    - 1075 turtlewater 
    - 1075 crocwater 
    - Celina turtle data
    - other d18Omw estimates from mid-K?

```{r, results='hide', message=FALSE, warning=FALSE}
# LATITUDINAL GRADIENTS ---------------------------------------------------

# Latitude vs MAT (see Table 1, Suarez et al. 2011)

# Leaf physiognomy Wolfe and Upchurch, 1987 <- - 0.0006 * lat2−0.2025* lat+30.25 
#LeafPhysCret <- -0.004* lat2+0.0796* lat−4.329 # Leaf physiognomy Wolfe and Upchurch, 1987; Spicer and Corfield, 1992


lat <- seq(0,90,by=1)
LatTempModern <- -0.003* (lat^2) - 0.2514 * lat+30.113 #Rozanski et al., 1993
LatTemp_CoolK  <- -0.004 * (lat^2) - 0.0586 * lat + 25.662 #Barron (1983)
LatTemp_WarmK <- -0.003 * (lat^2) + 0.0538 * lat + 28.534 #Barron (1983)
LatTemp_GENESIS_MOM <- -0.003 * (lat^2) - 0.0321 * lat + 32.514 #Zhou et al. (2008) 

LatGrads <- data.frame(lat, LatTempModern, LatTemp_CoolK, LatTemp_WarmK)
head(LatGrads)

#Kate's Antlers Data

AntPlat <- c(32, 32, 32, 32)
AntPhyTemps <- c(31,31,27,26)
AntlersPhyloTemps <- data.frame(AntPlat, AntPhyTemps)
AntlersPhyloTemps <- AntlersPhyloTemps %>% 
  rename(
    lat = AntPlat,
    temp = AntPhyTemps
  )

library(tidyverse)
df <- gather(LatGrads, key = ModelID, value = temp, 
       c("LatTempModern", "LatTemp_CoolK", "LatTemp_WarmK"))

# Suarez et al. 2021 Ruby Ranch Data

RRtemps <- c(38.6, 32.4, 31.3, 44.5, 19.8)
RRlat <- c(36.9, 36.9,36.9,36.9,36.9)
RRtempLat <- data.frame(RRtemps, RRlat)


# plot temp vs latitude with models overlain
plot(lat, LatTempModern, las = 1, type = "l", lwd = 2, ylim = c(-10, 40), xlab = substitute(paste(bold("Latitude (°N)"))), ylab=substitute(paste(bold("T (°C)"))), font = 2, cex.lab = 2, cex.axis = 1.5) 
box(lwd =3)
legend(0,15, 
       legend = c("Modern Gradient", "Cool Cretaceous Gradient", "Warm Cretaceous Gradient", "Cretaceous Climate Model"), 
       lty = c(1,2,3,5), lwd = 2,
       box.lwd = 0,
       bg ="transparent")
legend(0,0, 
       legend = c("Fish Phosphates (This Study)", "Phyllosilicates (Andrzejewski and Tabor, 2020)"), 
       pch = 1:2,
       box.lwd = 0,
       pt.cex = 1.5)
lines(LatTemp_CoolK, lty = 2, lwd = 2 )
lines(LatTemp_WarmK, lty = 3, lwd = 2)
lines(LatTemp_GENESIS_MOM, lty = 5, lwd = 2)
points(fishTempLat$lat, fishTempLat$temp, pch = 1, lwd = 2, cex = 1.5)
points(AntlersPhyloTemps$lat, AntlersPhyloTemps$temp, pch =2, lwd = 2, cex = 1.5)
#points(RRtempLat$RRlat, RRtempLat$RRtemps, pch =3, lwd = 2, cex = 1.5)

LatTemp_CoolK  <- -0.004 * (lat^2) - 0.0586 * lat + 25.662 #Barron (1983)
LatTemp_WarmK <- -0.003 * (lat^2) + 0.0538 * lat + 28.534 #Barron (1983)
LatTemp_GENESIS_MOM <- -0.003 * (lat^2) - 0.0321 * lat + 32.514 #Zhou et al. (2008) 

# Latitude vs Meteoric water d180 (see Table 1, Suarez et al. 2011)
LatMeteoric_Modern <- -0.003* (lat^2) + 0.0595* lat - 3.699 #Rozanski et al., 1993
LatMeteoric_CoolK <- -0.005 * (lat^2) + 0.1137 * lat - 5.270 #Barron (1983)
LatMeteoric_WarmK <- -0.005 * (lat^2) + 0.1299 * lat - 4.901 #Barron (1983)
LatMeteoric_GENESIS_MOM <- -0.005* (lat^2) +0.0730* lat - 4.001 #Zhou et al. (2008)




oxydeltmw <- expression("δ"^18 * "O‰ (VSMOW)")

par(mar = c(5,6,1,1))


# plot d18Omw vs latitude with models overlain
plot(lat, LatMeteoric_Modern, type = "l", las = 1, lwd = 2, xlim = c(20, 60), ylim = c(-10, -2), xlab = substitute(paste(bold("Latitude (°N)"))), ylab=substitute(paste(bold("d18Omw"))), font = 2, cex.lab = 2, cex.axis = 1.5) 
box(lwd =3)
legend("bottomleft", 
       box.lwd = 0,
       bg = "transparent",
       legend = c("Modern", "Cool K", "Warm K", "K Climate Model"), 
       lty = c(1,2,3,5), 
       lwd = 2 )
legend("topright",
       box.lwd = 0,
       bg = "transparent",
       legend = c("Turtles (Suarez et al., 2020)", "Glyptops sp.", "'Croc A'", "'Croc B'", "'Croc G'"), 
       pch = c(4, 7, 12, 10, 9),
       pt.cex = 1.5)
lines(LatMeteoric_CoolK, lty = 2, lwd = 2 )
lines(LatMeteoric_WarmK, lty = 3, lwd = 2)
lines(LatMeteoric_GENESIS_MOM, lty = 5, lwd = 2)
points(median(LSCelinaTurts$Palaeolatitude), median(LSCelinaTurts$d18Ow), pch = 4, col="black", lwd = 2, cex = 1.5)
points(median(HFCelinaTurts$Palaeolatitude), median(HFCelinaTurts$d18Ow), pch = 4, col="black", lwd = 2, cex = 1.5)
points(median(RRCelinaTurts$Palaeolatitude), median(RRCelinaTurts$d18Ow), pch = 4, col="black", lwd = 2, cex = 1.5)
points(c(40), d18Omw_turt, pch = 7, col="black", lwd = 2, cex = 1.5)
points(40, d18Ow_crocA, pch = 12, col="black", lwd = 2, cex = 1.5)
points(40, d18Omw_crocB, pch = 10, col="black", lwd = 2, cex = 1.5)
points(40, d18Omw_crocG, pch = 9, col="black", lwd = 2, cex = 1.5)
#points(RRCelinaTurts$Palaeolatitude, RRCelinaTurts$d18Ow, pch = 1, lwd = 2, cex = 1.5)
#points(HFCelinaTurts$Palaeolatitude, HFCelinaTurts$d18Ow, pch =2, lwd = 2, cex = 1.5)

dev.off()
```

