---
title: "1075 Stats"
author: "Matthew Allen"
date: "2023-11-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Evaluating sample size for multi-taxon bioapatite isotope studies

Accuracy and precision of paleo-temperature estimates derived from d18O of vertebrate bioapatite depend on intra-taxon sample size. That is, for each taxon or eco-type included in such an analysis, an adequate number of individuals from that group must be sampled in order to adequately reduce uncertainty and improve confidence in accuracy.

Multiple biological and environmental factors introduce variability into d18O distributions within taxonomic groups within a single fossil assemblage (i.e. "site"). The influence of each of these factors is likely to vary across time and space. Thus, it is necessary to evaluate such intra-taxon d18O variability within a single assemblage to understand what is an adequate sample size for each taxon in assemblages where it is represented.

Here I will demonstrate how sample size affects temperature estimates for a single assemblage in the Albian (mid-Cretaceous) of Montana. I will also demonstrate methods for determining target sample sizes for multi-taxon stable isotope analyses in general.

## Effect of intra-taxon d18Op sample size on temperature estimates

```{r}
# Ultimately this code will source data from Github. For now it is local

# Here I will include a code chunk that will:
  # load the required packages
  # source the data (V1075_BySpec.csv)
  # plot a histogram for each eco_type to visualize the d18Op distributions
 
```

## Histograms of d18Op for each eco_type (taxon)

Here will be a plot as described. I will include observations about noteworthy distributions here (e.g. note that small theropod distribution is bimodal, fish distribution is normal, etc.).

```{r pressure, echo=TRUE}
plot(pressure)
```

## Compute proxy parameters. Propagate uncertainties or run sensitivity analyses.

```{r}
 # Here I will include code chunk that will:
  # Compute d18Omw and temp proxy estimates with uncertainties
  # 
     
```

## Demonstrate direct Monte Carlo simulation and boostrapping as methods to determine target sample size.

```{r}
# Here I will include code chunk that will:
  # run direct Monte Carlo simulation of 1000 d18Op values
  # run 10,000 bootstraps on simulated data
  # generate plots that show 
    # relationship between sample size and precision of means
    # relationships between sample size and estimated accuracy of means (95% confidence that true mean was captured by   each subsample)
```

```{r}
# PLACEHOLDER

```
