---
title: "midK Latitudinal Gradients"
author: "Matthew Allen"
date: "2024-01-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
oxydeltmw <- expression("δ"^18 * "O‰ (VSMOW)")
```

## Setup


```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)


 # read in V1075_MCwater.csv (LOCAL)
  setwd("/Users/allen/Documents/GitHub/1075_Vertebrate_d18Op/Data")
  V1075_MCwater <- read.csv("V1075_MCwater.csv")


  # Latitude vs MAT (see Table 1, Suarez et al. 2011)

    # Leaf physiognomy Wolfe and Upchurch, 1987 <- - 0.0006 * lat2−0.2025* lat+30.25 
    #LeafPhysCret <- -0.004* lat2+0.0796* lat−4.329 # Leaf physiognomy Wolfe and Upchurch, 1987; Spicer and Corfield, 1992
    
    # Empirical Data
      # Temps from V1075 dual-taxon results
        TempFishTurt <- 22.2 
        TemplowCI <- 22.2 - 5.5
        TemphighCI <- 22.2 + 5.4
        
      # V1075 d18O water data
        # Glyptops
          turtle_1075 <- -8.4
          turtle_1075_loCI <- turtle_1075 - 1.1
          turtle_1075_hiCI <- turtle_1075 +1.1
        # Croc G 
          # CrocG_1075 <- 
          # CrocG_1075_loCI <-
          # CrocG_1075_hiCI <-
          
      # Suarez et al. 2021 Ruby Ranch Data
          
        # Ruby Ranch paleolats
          RRR_paleolats <- c(53-6, 47-6, 51-6, 50-6) # See 4 model options at paleolatitude.org, based on 6° difference from modern Cloverly lat/long of 45°N/-108°W
          RRlat <- mean(RRR_paleolats)
          
          # create new dataframe with V1075 temp datapoint and paleolat
          RRtemps <- c(38.6, 32.4, 31.3, 44.5, 19.8)
          RRTempLat <- data.frame(cbind(temp = RRtemps, lat = RRlat, proxy = "D47 CO3"))
          

        
      #Kate's Antlers Data
          # Trinity paleolats
          OKTX_paleolats <- c(53-8, 47-8, 51-8, 50-8) # See 4 model options at paleolatitude.org, based on 6° difference from modern Cloverly lat/long of 45°N/-108°W
          OKTXlat <- mean(OKTX_paleolats)

        # create new dataframe with V1075 temp datapoint and paleolat
        OKTX_Temps <- c(31,31,27,26)
        OKTX_TempLat <- data.frame(cbind(temp = OKTX_Temps, lat = OKTXlat, proxy = "D47 CO3"))
        
    # Models
      lat <- seq(0,90,by=1)
      LatTempModern <- -0.003* (lat^2) - 0.2514 * lat+30.113 #Rozanski et al., 1993
      LatTemp_CoolK  <- -0.004 * (lat^2) - 0.0586 * lat + 25.662 #Barron (1983)
      LatTemp_WarmK <- -0.003 * (lat^2) + 0.0538 * lat + 28.534 #Barron (1983)
      LatTemp_GENESIS_MOM <- -0.003 * (lat^2) - 0.0321 * lat + 32.514 #Zhou et al. (2008) 
      
      LatGrads <- data.frame(lat, LatTempModern, LatTemp_GENESIS_MOM)
      head(LatGrads)


      df <- gather(LatGrads, key = ModelID, value = temp, 
             c("LatTempModern", "LatTemp_GENESIS_MOM"))



df$ModelID <- as.factor(df$ModelID)
levels(df$ModelID)
df$ModelID <- factor(df$ModelID, levels = c("LatTempModern", "LatTemp_CoolK", "LatTemp_WarmK"))

# create new dataframe with V1075 temp datapoint and paleolat
Cloverly_paleolats <- c(53, 47, 51, 50) # See 4 model options at paleolatitude.org, based on modern lat/long of 45°N/-108°W
l <- mean(Cloverly_paleolats)
V1075TempLat <- data.frame(cbind(temp = c(TempFishTurt), lat = l, ModelID = "gar"))

```

## Plot Temps


```{r, echo=FALSE}
ggarV1075 <- V1075TempLat$temp
par(mar = c(12,12,1,10))

plot(lat, LatTempModern, 
     las = 1, type = "l", 
     lwd = 1, 
     ylim = c(-10, 40), 
     xlim = c(30,60), 
     xlab = substitute(paste("Latitude (°N)")), 
     ylab=substitute(paste("T (°C)")), 
     font = 1, 
     cex.lab = 1, 
     cex.axis = 1) 

#box(lwd =3)
legend("topright",
       inset=c(-0.75, 0),
       #xpd = "TRUE",
       box.lwd = 0,
       legend = c("Modern Gradient", "GENESIS-MOM"), 
       lty = c(1,2,3,5), lwd = 2,
       bg ="transparent")
legend("bottomright", 
       inset=c(-0.75, 0),
       xpd = "TRUE",
       legend = c("Fossil Phosphate 
(This Study)", "Phyllosilicates", "D47 CO3"), 
       pch = 1:2:3,
       box.lwd = 0,
       pt.cex = 1.5)
#lines(LatTemp_CoolK, lty = 2, lwd = 2 )
#lines(LatTemp_WarmK, lty = 3, lwd = 2)
lines(LatTemp_GENESIS_MOM, lty = 5, lwd = 2)
points(V1075TempLat$lat, V1075TempLat$temp, pch = 1, lwd = 2, cex = 1)
points(OKTX_TempLat$lat, OKTX_TempLat$temp, pch =2, lwd = 2, cex = 1)
points(RRTempLat$lat, RRTempLat$temp, pch =3, lwd = 2, cex = 1)


```

## Plot Water Isotopes


```{r, echo=FALSE}
# Aptian-Albian turtles (Suarez et al., 2020)

    setwd("/Users/allen/Documents/Data Analysis/Data/Geochem")
    CelinaTurtles <- read.csv("SuarezEtAl2020_AptianAlbianTurtleDATA.csv")
    
    table(CelinaTurtles$Formation)
    table(CelinaTurtles$Member)
    
    RRCelinaTurts <- CelinaTurtles[which(CelinaTurtles$Member %in% c("Ruby Ranch")), ]
    LSCelinaTurts <- CelinaTurtles[which(CelinaTurtles$Member %in% c("Little Sheep")), ]
    HFCelinaTurts <- CelinaTurtles[which(CelinaTurtles$Member %in% c("")), ]

    # updated paleolat model
    RRCelinaTurts$Palaeolatitude <- RRlat
    
    # Holly Fork 
    HFlat <- mean(c(53-10.5, 47-10.5, 51-10.5, 50-10.5))
    HFCelinaTurts$Palaeolatitude <- HFlat
    
    # Cloverly 
    LSCelinaTurts$Palaeolatitude <- l



# Latitude vs Meteoric water d180 (see Table 1, Suarez et al. 2011)
LatMeteoric_Modern <- -0.003* (lat^2) + 0.0595* lat - 3.699 #Rozanski et al., 1993
LatMeteoric_CoolK <- -0.005 * (lat^2) + 0.1137 * lat - 5.270 #Barron (1983)
LatMeteoric_WarmK <- -0.005 * (lat^2) + 0.1299 * lat - 4.901 #Barron (1983)
LatMeteoric_GENESIS_MOM <- -0.005* (lat^2) +0.0730* lat - 4.001 #Zhou et al. (2008)

LatPedocarb <- -0.0042* (lat^2) + 0.08* lat - 4.33 #Suarez et al. 2011







# create delta notation for label
oxydeltwater <- expression(paste(delta^{18}, "O"[water], " (‰ V-SMOW)"))


par(mar = c(5,5,1,10))

# plot d18Omw vs latitude with models overlain
plot(lat, 
     LatMeteoric_Modern,
     type = "l",
     las = 1,
     lwd = 2,
     xlim = c(30, 60),
     ylim = c(-10, -2),
     xlab = substitute(paste("Latitude (°N)")),
     ylab=(oxydeltwater),
     font = 1,
     cex.lab = 1,
     cex.axis = 1) 

#box(lwd =3)
legend("bottomright",
       inset=c(-0.5, 0),
       box.lwd = 0,
       bg = "transparent",
       legend = c("Modern", "GEN-MOM", "Siderites"), 
       lty = c(1,3,4),
       xpd = TRUE,
       lwd = 2 )


legend("topright",
       inset=c(-0.5, 0),
       xpd = TRUE,
       box.lwd = 0,
       bg = "transparent",
       legend = c("Glyptops", "'Croc A'", "'Croc B'", "'Croc G'", "Naomichelys", "Celina's Turtles"), 
       pch = c(7, 12, 10, 9, 8, 4),
       pt.cex = 1.5)
#lines(LatMeteoric_CoolK, lty = 2, lwd = 2 )
#lines(LatMeteoric_WarmK, lty = 3, lwd = 2)
lines(LatMeteoric_GENESIS_MOM, lty = 3, lwd = 2)
lines(LatPedocarb, lty = 4, lwd = 2)
points(median(LSCelinaTurts$Palaeolatitude), median(LSCelinaTurts$d18Ow), pch = 4, col="black", lwd = 1, cex = 1)
points(median(HFCelinaTurts$Palaeolatitude), median(HFCelinaTurts$d18Ow), pch = 4, col="black", lwd = 1, cex = 1)
points(median(RRCelinaTurts$Palaeolatitude), median(RRCelinaTurts$d18Ow), pch = 4, col="black", lwd = 1, cex = 1)
points(c(50.25), V1075_MCwater$MEAN[which(V1075_MCwater$Taxon == "Aquatic Turtle")], pch = 7, col="black", lwd = 1, cex = 1)
points(50.25, V1075_MCwater$MEAN[which(V1075_MCwater$Taxon == "Croc A")], pch = 12, col="black", lwd = 1, cex = 1)
points(50.25, V1075_MCwater$MEAN[which(V1075_MCwater$Taxon == "Croc B")], pch = 10, col="black", lwd = 1, cex = 1)
points(50.25, V1075_MCwater$MEAN[which(V1075_MCwater$Taxon == "Croc G")], pch = 9, col="black", lwd = 1, cex = 1)
points(50.25, V1075_MCwater$MEAN[which(V1075_MCwater$Taxon == "Terrestrial Turtle")], pch = 8, col="black", lwd = 1, cex = 1)
