---
title: "Monte Carlo Simulation of d18O Estimates"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
# Load required libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(knitr)

# Set global chunk options
knitr::opts_chunk$set(echo = FALSE)
```

# Monte Carlo Simulation of \( \delta^{18}O \) Estimates

## 1. Data Preparation

```{r data-preparation}
# Load input distributions (replace paths with actual file locations)
load("~/Documents/GitHub/1075_Vertebrate_d18Op/Data/synth_NIST.RData")
load("~/Documents/GitHub/1075_Vertebrate_d18Op/Data/MC_d18Op_dist.RData")
# Extract d18Op data for Glyptops
d18Op_synthglyp <- MC_d18Op_dist$values[MC_d18Op_dist$group == "Glyptops"]

# Load regression models
Barrick_lm_model <- readRDS("path/to/Barrick_regression_model.rds")
Puceat_lm_model <- readRDS("path/to/Puceat_regression_model.rds")
```

---

## 2. Barrick Regression Analysis

### Model Summary

```{r barrick-model-summary}
# Extract model statistics
summary_barrick <- summary(Barrick_lm_model)
intercept_barrick <- coef(summary_barrick)[1, "Estimate"]
slope_barrick <- coef(summary_barrick)[2, "Estimate"]
intercept_se_barrick <- coef(summary_barrick)[1, "Std. Error"]
slope_se_barrick <- coef(summary_barrick)[2, "Std. Error"]
residual_sd_barrick <- summary_barrick$sigma
```

- **Intercept:** `r round(intercept_barrick, 2)` ± `r round(intercept_se_barrick, 2)`
- **Slope:** `r round(slope_barrick, 2)` ± `r round(slope_se_barrick, 2)`
- **Residual SD:** `r round(residual_sd_barrick, 2)`

---

## 3. Monte Carlo Simulation for Water \( \delta^{18}O \)

```{r monte-carlo-water}
# Set up simulation parameters
n_iterations <- 10000
set.seed(123)

# Simulate regression coefficients
simulated_slope <- rnorm(n_iterations, mean = slope_barrick, sd = slope_se_barrick)
simulated_intercept <- rnorm(n_iterations, mean = intercept_barrick, sd = intercept_se_barrick)

# Simulate residual error
residual_error <- rnorm(n_iterations, mean = 0, sd = residual_sd_barrick)

# Perform Monte Carlo simulation
water_simulations <- outer(d18Op_synthglyp, simulated_slope, "*") + simulated_intercept + residual_error
```

### Results

```{r water-results}
# Compute statistics
water_means <- rowMeans(water_simulations)
water_sd <- apply(water_simulations, 1, sd)
water_ci <- apply(water_simulations, 1, function(x) quantile(x, probs = c(0.025, 0.975)))

# Extract overall statistics
mean_water <- mean(water_means)
lower_95_CI_water <- min(water_ci[1, ])
upper_95_CI_water <- max(water_ci[2, ])

# Visualize
ggplot(data = data.frame(Simulations = water_means), aes(x = Simulations)) +
  geom_histogram(bins = 50, fill = "blue", alpha = 0.7) +
  geom_vline(xintercept = c(lower_95_CI_water, upper_95_CI_water), linetype = "dashed", color = "red") +
  labs(
    title = "Distribution of Simulated Water \( \delta^{18}O \) Estimates",
    x = "Water \( \delta^{18}O \) (‰)",
    y = "Frequency"
  )
```

- **Mean Water \( \delta^{18}O \):** `r round(mean_water, 2)`
- **95% Confidence Interval:** [`r round(lower_95_CI_water, 2)`, `r round(upper_95_CI_water, 2)`]

---

## 4. Puceat Regression Analysis

### Model Summary

```{r puceat-model-summary}
# Extract model statistics
summary_puceat <- summary(Puceat_lm_model)
intercept_puceat <- coef(summary_puceat)[1, "Estimate"]
slope_puceat <- coef(summary_puceat)[2, "Estimate"]
intercept_se_puceat <- coef(summary_puceat)[1, "Std. Error"]
slope_se_puceat <- coef(summary_puceat)[2, "Std. Error"]
residual_sd_puceat <- summary_puceat$sigma
```

- **Intercept:** `r round(intercept_puceat, 2)` ± `r round(intercept_se_puceat, 2)`
- **Slope:** `r round(slope_puceat, 2)` ± `r round(slope_se_puceat, 2)`
- **Residual SD:** `r round(residual_sd_puceat, 2)`

---

## 5. Monte Carlo Simulation for Temperature

```{r monte-carlo-temp}
# Simulate temperature predictions
temperature_simulations <- outer(water_means, simulated_slope, "*") + simulated_intercept + residual_error
temp_means <- rowMeans(temperature_simulations)
temp_sd <- apply(temperature_simulations, 1, sd)
temp_ci <- apply(temperature_simulations, 1, function(x) quantile(x, probs = c(0.025, 0.975)))

# Extract overall statistics
mean_temp <- mean(temp_means)
lower_95_CI_temp <- min(temp_ci[1, ])
upper_95_CI_temp <- max(temp_ci[2, ])

# Visualize
ggplot(data = data.frame(Simulations = temp_means), aes(x = Simulations)) +
  geom_histogram(bins = 50, fill = "orange", alpha = 0.7) +
  geom_vline(xintercept = c(lower_95_CI_temp, upper_95_CI_temp), linetype = "dashed", color = "red") +
  labs(
    title = "Distribution of Simulated Temperature Estimates",
    x = "Temperature (°C)",
    y = "Frequency"
  )
```

- **Mean Temperature:** `r round(mean_temp, 2) °C`
- **95% Confidence Interval:** [`r round(lower_95_CI_temp, 2)`, `r round(upper_95_CI_temp, 2)`]

---

## 6. Error Contributions

```{r error-contributions}
# Baseline standard deviation
baseline_sd <- sd(temp_means)

# Fixed variable contributions
delta_Op_fixed <- mean(delta_Op_distribution)
delta_Ow_fixed <- mean(water_means)

# Compute contributions
contributions <- c(
  delta_Op = sd(outer(delta_Op_fixed, simulated_slope, "*") + simulated_intercept + residual_error),
  delta_Ow = sd(outer(delta_Ow_fixed, simulated_slope, "*") + simulated_intercept + residual_error)
)

contributions <- contributions / baseline_sd * 100

kable(data.frame(
  Variable = names(contributions),
  Contribution = round(contributions, 2)
))
